---
title: "Assignment_1_Cereals"
author: "Shujath"
date: "2025-09-07"
output:
  html_document: default
  pdf_document: default
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


**Load Libraries - Intsalling Packages**
```{r}
library(tidyverse)
```


## Task_1_&_2: Downloading and Importing Dataset of Cereals from CSV Files of Module-1 

```{r}
library(readr)
Cereals <- read_csv("~/Desktop/Fundamentals of Machine Learning - Dr. Mostafa Kamali/Cereals.csv")
head(Cereals)
```


## Task_3: Printing out descriptive statistics for a selection of quantitative and categorical variables 

**Data Cleaning - Identifying the number of missing values in columns**
```{r}
colSums(is.na(Cereals)) 
```
**Replacing the missing values with 0**
```{r}
#Replacing missing values with 0 in selected and identified columns
Cereals$carbo[is.na(Cereals$carbo)] <- 0
Cereals$sugars[is.na(Cereals$sugars)] <- 0
Cereals$potass[is.na(Cereals$potass)] <- 0

#Re-running a check 
colSums(is.na(Cereals))
```


## Task_4: Transformaing atleast one variable - Creating a new column to understand the protein to calorie ratio

```{r}
Cereals <- Cereals %>%
  mutate(Protein_per_Calorie = protein / calories)
summary(Cereals)
```


## Task_5: Plotting atleast one quantitaive variable, and one scatterplot

**Histogram**
```{r}
ggplot(Cereals, aes(x = calories)) +
  geom_histogram(binwidth = 10, fill = "steelblue", color = "black") +
  labs(title = "Distribution of Calories in Cereals",
       x = "Calories per Serving",
       y = "Count of Cereals")
```

**Scatterplot**
```{r}
ggplot(Cereals, aes(x = sugars, y = rating, color = type)) +
  geom_point(size = 3, alpha = 0.7) +
  labs(title = "Relationship between Sugars and Rating",
       x = "Sugar Content (grams)",
       y = "Cereal Rating")
```

